---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# lazymatrix

<!-- badges: start -->
[![R-CMD-check](https://github.com/vsegersall/lazymatrix/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/vsegersall/lazymatrix/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`lazymatrix` provides a framework for working with transformed sparse data matrices using [lazy evaluation](https://en.wikipedia.org/wiki/Lazy_evaluation). This approach is particularly useful for handling large sparse matrices where transformations (e.g., centering and scaling) would otherwise result in dense matrices, consuming significant memory and computational resources.

## Installation

You can install the development version of lazymatrix from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("vsegersall/lazymatrix")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(lazymatrix)
set.seed(123)
sparse_matrix <- Matrix::Matrix(0, 5, 3)
sparse_matrix[sample(length(sparse_matrix), 5)] <- rnorm(5)
b <- rnorm(3)
lazy_matrix <- LazyMatrix(sparse_matrix, scale = "sd", location = "mean")
print(lazy_matrix)
lazy_matrix %*% b
```

## Contribution
As this is part of a university project and the fact that the project does not have a stable version yet, the project is closed for contributions. We are however happy to discuss ideas for the future so feel free to contact the authors or make an issue for functionalities you would like to see within `lazymatrix`.