---
title: "introduction_lazymatrix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction_lazymatrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## 1. Why LazyMatrix?
When working with sparse matrices in R, a challenge is performing transformations on the matrix without loosing sparsity. An operation such as dividing the elements with it's sample variance and subtracting the sample mean removes sparsity and risks occupying a lot of memory. As memory often is limited, creating copies of the transformed matrix, which is no longer sparse, may not be desirable or even feasible. This creates issues when we aim to work with a scaled version of the design matrix for statistical algorithms such as principal component analysis or linear regression. A solution is working with the `LazyMatrix`-object, which never stores copies of the transformed matrix but performs operations only when neccessary for the desired algorithm.

## 2. Mathematical Background
Assume we want to perform the  matrix operation

$$
\tilde{A}b
$$

where $\tilde{A}$ is a scaled version of the matrix $A$ being $n\times p$, and the vector $b\in R^p$. Assume furhter that we want to transform $A$ subtracting some location parameter $C$ and dividing by some scale $S$. This operation is defined as

$$
(A-C)S^{-1}b.
$$

Now, assuming that $A$ is sparse, subtracting $C$ will result in a non-sparse matrix as all zero, or close to zero, elements will now be non-zero if $c_i\neq0$. The simple solution is basically expanding the expression so that we get
$$
(A-C)S^{-1}b=AS^{-1}b-CS^{-1}b
$$

hence an equivalent formalization where we avoid materializing $A$. This reasoning can be applied to all matrix operations but we will see that just a few of them are neccessary to perform complex statistical algorithms.

## 3. Usage
The way of doing this in R is by creating the S4 class `LazyMatrix` which stores the original data matrix, location and scale parameters.
```{r setup}
library(lazymatrix)
set.seed(123)
sparse_matrix <- Matrix::Matrix(0, 5, 3)
sparse_matrix[sample(length(sparse_matrix), 5)] <- rnorm(5)
b <- rnorm(3)
lazy_matrix <- LazyMatrix(sparse_matrix, scale = "sd", location = "mean")
lazy_matrix
```
Mainly we are interested in the location and scale parameters of the columns, as these usually represents the variables. However, storing the location and scale for the rows as well allows for working with the transpose of the matrix by just inverting the rows and columns.
```{r transpose}
lazy_transpose <- t(lazy_matrix)
lazy_transpose
```
As we aim for `LazyMatrix` to be as user-friendly as possible, the operations are defined as is standard in R. Functions from `base` such as `nrow`, `ncol` and `colnames` are designed to work equivalently. For matrix operations, we design them to be as straight forward from the normal syntax as possible. Below follow the examples of $\tilde{A}b$ or $\tilde{A}^Tc$, where $b\in R^p$ and $c\in R^n$.

```{r matrix operations}
set.seed(123)
b <- rnorm(ncol(lazy_matrix))
c <- rnorm(nrow(lazy_matrix))
product <- lazy_matrix %*% b
cross_product <- crossprod(lazy_matrix, c)
product
cross_product
```

## 4. Design
This package has been designed carefully to align with the framework `Matrix`. As this package already has efficient ways of handling sparse matrices using `sparseMatrix`, `lazymatrix` should be seen as an extension which allows for performing efficient lazy computations.