---
title: "linear_regression"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{linear_regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# 1. Linear Regression
## 1.1. Introduction
Several algorithms for iterative least squares algorithms have been proposed when working with sparse matrices. We have chosen to work with the method proposed by Paige and Saunders (1982) called LSQR [@paige1982lsqr]. This iterative algorithm allows us to use custom methods for matrix operations which allows for our lazy computations. We will see that the operations `%*%`, `crossprod` and `norm` are enough for getting stable estimates of the regression coefficients.

## 1.2. Notation
The authors proposes an algorithm as a set of initalisation parameters and the design matrix $A$. As mentioned by the authors, the matrix $A$ is used only to compute products of the form $Av$ and $A^Tu$. This object will be our `LazyMatrix` objects and hence, all computations will be done lazily at every step of the iteration. For clarity, matrices will be denoted by capital letters $A, B,\cdots$, vectors by $v,w,\cdots$ and scalars by greek letters $\alpha, \beta,\cdots$. For vectors, the norm is the Euclidean norm and for matrices we assume the Frobenius norm $||A||_F=\sqrt{\sum |a_{ij}|^2}$.

## 1.3. Mathematical Background
We will make use of the [Lanczos process](https://en.wikipedia.org/wiki/Lanczos_algorithm), for which we will not go into further detail. The main idea is that with the help of a set of initialization parameters, we can with the help of the design matrix $A$ update the regression coefficients for every step of the algorithm. The algorithm is explained as

1. Initialization:
   \begin{align*}
      \beta_1 &= \sqrt{\sum b^2}, \\
      \mathbf{u}_1 &= \frac{\mathbf{b}}{\beta_1}, \\
      \mathbf{A}^\top \mathbf{u}_1 &= \mathbf{A}^\top \mathbf{u}_1, \\
      \alpha_1 &= \sqrt{\sum (\mathbf{A}^\top \mathbf{u}_1)^2}, \\
      \mathbf{v}_1 &= \frac{\mathbf{A}^\top \mathbf{u}_1}{\alpha_1}, \\
      \mathbf{w}_1 &= \mathbf{v}_1, \\
      \mathbf{x}_0 &= \mathbf{0}, \\
      \bar{\phi_1} &= \beta_1, \\
      \bar{\rho_1} &= \alpha_1.
   \end{align*}

2. For $i = 1, 2, \dots$, repeat steps 3--6:

   3. Continue the bidiagonalization:
   \begin{align*}
      \mathbf{\beta}_u &= \mathbf{A} \mathbf{v}_1 - \alpha_1 \mathbf{u}_1, \\
      \beta_2 &= \sqrt{\sum \mathbf{\beta}_u^2}, \\
      \mathbf{u}_2 &= \frac{\mathbf{\beta}_u}{\beta_2}, \\
      \mathbf{\alpha}_v &= \mathbf{A}^\top \mathbf{u}_2 - \beta_2 \mathbf{v}_1, \\
      \alpha_2 &= \sqrt{\sum \mathbf{\alpha}_v^2}, \\
      \mathbf{v}_2 &= \frac{\mathbf{\alpha}_v}{\alpha_2}.
   \end{align*}

   4. Construct and apply the next orthogonal transformation:
   \begin{align*}
      \rho_1 &= \sqrt{(\bar{\rho_1})^2 + \beta_2^2}, \\
      c_1 &= \frac{\bar{\rho_1}}{\rho_1}, \\
      s_1 &= \frac{\beta_2}{\rho_1}, \\
      \theta_2 &= s_1 \alpha_2, \\
      \bar{\rho_2} &= -c_1 \alpha_2, \\
      \phi_1 &= c_1\bar{\phi_1}, \\
      \bar{\phi_2} &= s_1 \bar{\phi_1}.
   \end{align*}

   5. Update $\mathbf{x}$ and $\mathbf{w}$:
   \begin{align*}
      \mathbf{x}_1 &= \mathbf{x}_0 + \frac{\phi_1}{\rho_1} \mathbf{w}_1, \\
      \mathbf{w}_2 &= \mathbf{v}_2 - \frac{\theta_2}{\rho_1} \mathbf{w}_1.
   \end{align*}

   6. Reset the loop variables:
   \begin{align*}
      \beta_1 &= \beta_2, \\
      \mathbf{u}_1 &= \mathbf{u}_2, \\
      \alpha_1 &= \alpha_2, \\
      \mathbf{v}_1 &= \mathbf{v}_2, \\
      \bar{\rho_1} &= \bar{\rho_2}, \\
      \bar{\phi_1} &= \bar{\phi_2}, \\
      \mathbf{x}_0 &= \mathbf{x}_1, \\
      \mathbf{w}_1 &= \mathbf{w}_2.
   \end{align*}

   7. Check for convergence.

At this point, the authors do not propose a criterion for when we should stop iterating. However, we have chosen to use the criterion that [LSQR in scipy](https://docs.scipy.org/doc/scipy/reference/generated/scipy.sparse.linalg.lsqr.html) [@virtanen2020scipy; @scipy_lsqr] uses. First, compute the residual between the response vector $y$ and the estimate as

$$
r = y-Ax_0
$$

and then we stop iterating when

$$
\|r\| \leq \text{tol}\cdot \|A\| \cdot \|x_0\| + \text{tol} \cdot \|b\|.
$$

`tol` is a parameter for tolerance level of the difference that we accept. In our algorithm, it is by default $1e-6$.

## 1.4. Implementation
Assuming we want to fit regression coefficient vector $\beta$ on the scaled matrix $X$ with response vector $y$. We can compute these estimates using lazy computation. Assuming the original sparse matrix `A`, create an instance of `LazyMatrix` and run the method `lazymatrix::lsqr`.
```{r setup}
library(lazymatrix)
 # 1. Define sparseMatrix
set.seed(123)
n_row <- 50
n_col <- 10
i <- c(
  1:50,
  sample(1:50, 20, replace = TRUE),
  sample(1:50, 15, replace = TRUE)
)
j <- c(
  rep_len(1:10, 50),
  sample(1:10, 20, replace = TRUE),
  sample(1:10, 15, replace = TRUE)
)
pairs <- unique(data.frame(i = i, j = j))
i <- pairs$i
j <- pairs$j
x <- rnorm(length(i))
A <- Matrix::sparseMatrix(i = i, j = j, x = x, dims = c(n_row, n_col))

# 2. Define response vector
y <- rnorm(nrow(A))

# 3. Define LazyMatrix
X <- LazyMatrix(A, "sd", "mean")

# 4. Perform lsqr
lazy_beta <- lsqr(X, y)
```
Now,the non-lazy approach would be to scale `A` using `base::scale`and then use something like `stats::lm.fit` to get the estimates. This would imply creating a copy of `A` with non-zero elements taking up unneccessary memory and being much slower generally.
```{r non lazy approach}
scaled_a <- base::scale(A)
non_lazy_beta <- stats::lm.fit(scaled_a, y)$coefficients
```
The estimates are equivalent as is shown below.
```{r result}
isTRUE(all.equal(as.vector(lazy_beta), as.vector(non_lazy_beta)))
```
# 2. Gradient Descent
## 2.1. Introduction
In modern statistical modeling, the gradient descent algorithm is fundamental for any kind of modern machine learning approach. The approach we propose based on lazy evaluation, is how to perform the simplest form of gradient descent on a scaled matrix. As many of our implementations show, it is sufficient to use a small set of matrix operations in order to perform this algorithm.

## 2.2. Mathematical Background
The algorithm follows the approach as explained by @ng2025cs229. Assume the cost function $J(\theta)$ defined as

$$
J(\theta)=\frac{1}{2}\sum_{i=1}^n(h_\theta(x^{(i)})-y^{(i)})^2
$$

where $h_\theta(x)=\theta^Tx$, $x$ being the elements of the design matrix and $\theta$ the parameters. In gradient descent, we usually talk about two types of parameters, weights and biases. In the notation above, they are both included within the parameter vector $\theta$. This is done by letting $x_0=1$ so that the first term in $\sum_{i=0}^n\theta_ix_i$ includes the additive bias term. $y$ is the response vector, so we recognize $J$ to be the cost function representing the residual sum of squares. Hence, the goal is to choose $\theta$ to minimize $J$. We start with an inistial guess of the parameter and repeatedly update it to make the cost function smaller. This is done with the gradient descent algorithm defined as

$$
\theta_j:=\theta_j - \alpha\frac{\partial}{\partial\theta_j}J(\theta)
$$
where the partial derivative is the derivative of the cost function with regards to the parameter. $\alpha$ is the learning rate that usually takes a small value to control how fast the algorithm converges. Computing this derivative is straight forward and we get the expression

$$
\frac{\partial}{\partial\theta_j}J(\theta)=(h_\theta(x)-y)x_j
$$

providing us with the update rule

$$
\theta_j:=\theta_j+\alpha(y^{(i)}-h_\theta(x^{(i)}))x_j^{(i)}.
$$

## 2.3. Implementation
The following gradient descent algorithm works as helper function for which we will compare the lazy-approach to the non-lazy.
```{r gradient descent algorithm}
gradient_descent_helper <- function(
  x,
  y,
  w_init,
  b_init,
  learning_rate,
  n_epochs
) {
  w <- w_init
  b <- Matrix::Matrix(b_init, nrow = nrow(x))
  n <- nrow(x)
  for (epoch in 1:n_epochs) {
    y_pred <- x %*% w + b
    error <- y_pred - y
    w <- w - (learning_rate * crossprod(x, error)) / n
    b <- b - (learning_rate * sum(error)) / n
  }
  return(list(w = w, b = b))
}
```
Compared to the definition above, in this algorithm we include the bias term as `b`. This results a list of the two parameter vectors, weights and biases. `epochs` is the number of time we choose to iterate. As the result of `LazyMatrix %*% vector` results in a `Matrix`-object, the addition of the bias term can be handled completely within the `Matrix` framework. The key of `lazymatrix` is to handle the heavy matrix operations lazily which is done in this case using `%*%` and `crossprod`. Now, we use the sparse matrix `A` from section #1 and it's `LazyMatrix` `X` and set initial values randomly for the weights and biases and the response vector `y_true`.

```{r initialization}
  set.seed(4567)
  p <- ncol(X)
  w_init <- stats::rnorm(p)
  b_init <- stats::rnorm(1)
  y_true <- stats::rnorm(nrow(A))
  learning_rate <- 0.01
  n_epochs <- 10
```
Note that the function returns the parameters from the iterative algorithm.
```{r}
pars_lazy <- gradient_descent_helper(
   x = X,
   y = y_true,
   w_init = w_init,
   b_init = b_init,
   learning_rate = learning_rate,
   n_epochs = n_epochs
  )
print("First 5 parameter weights and the bias term.")
print(pars_lazy$w[1:5])
print(pars_lazy$b[1])
```
These parameters can now be used for predictions and are comparible with the non-lazy way of using `scale(A)`and then running gradient descent.

## 2.4. Discussion
Nowadays there exists much more complex versions of the gradient descent algorithm, but this example proves that just with a few matrix operations being performed lazily, we can avoid storing heavy copies of large matrices and only perform operations when they are needed. The idea is to implement methods to `lazymatrix` which allows for more complex algorithms such as Stochastic gradient descent.

# 4. References
