---
title: "linear_regression"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{linear_regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# 1. Linear Regression
## 1.1. Introduction
Several algorithms for iterative least squares algorithms have been proposed when working with sparse matrices. We have chosen to work with the method proposed by Paige and Saunders (1982) called LSQR [@paige1982lsqr]. This iterative algorithm allows us to use custom methods for matrix operations which allows for our lazy computations. We will see that the operations `%*%`, `crossprod` and `norm` are enough for getting stable estimates of the regression coefficients.

## 1.2. Notation
The authors proposes an algorithm as a set of initalisation parameters and the design matrix $A$. As mentioned by the authors, the matrix $A$ is used only to compute products of the form $Av$ and $A^Tu$. This object will be our `LazyMatrix` objects and hence, all computations will be done lazily at every step of the iteration. For clarity, matrices will be denoted by capital letters $A, B,\cdots$, vectors by $v,w,\cdots$ and scalars by greek letters $\alpha, \beta,\cdots$. For vectors, the norm is the Euclidean norm and for matrices we assume the Frobenius norm $||A||_F=\sqrt{\sum |a_{ij}|^2}$.

## 1.3. Mathematical Background
We will make use of the [Lanczos process](https://en.wikipedia.org/wiki/Lanczos_algorithm), for which we will not go into further detail. The main idea is that with the help of a set of initialization parameters, we can with the help of the design matrix $A$ update the regression coefficients for every step of the algorithm. The algorithm is explained as

1. Initialization:
   \begin{align*}
      \beta_1 &= \sqrt{\sum b^2}, \\
      \mathbf{u}_1 &= \frac{\mathbf{b}}{\beta_1}, \\
      \mathbf{A}^\top \mathbf{u}_1 &= \mathbf{A}^\top \mathbf{u}_1, \\
      \alpha_1 &= \sqrt{\sum (\mathbf{A}^\top \mathbf{u}_1)^2}, \\
      \mathbf{v}_1 &= \frac{\mathbf{A}^\top \mathbf{u}_1}{\alpha_1}, \\
      \mathbf{w}_1 &= \mathbf{v}_1, \\
      \mathbf{x}_0 &= \mathbf{0}, \\
      \bar{\phi_1} &= \beta_1, \\
      \bar{\rho_1} &= \alpha_1.
   \end{align*}

2. For $i = 1, 2, \dots$, repeat steps 3--6:

   3. Continue the bidiagonalization:
   \begin{align*}
      \mathbf{\beta}_u &= \mathbf{A} \mathbf{v}_1 - \alpha_1 \mathbf{u}_1, \\
      \beta_2 &= \sqrt{\sum \mathbf{\beta}_u^2}, \\
      \mathbf{u}_2 &= \frac{\mathbf{\beta}_u}{\beta_2}, \\
      \mathbf{\alpha}_v &= \mathbf{A}^\top \mathbf{u}_2 - \beta_2 \mathbf{v}_1, \\
      \alpha_2 &= \sqrt{\sum \mathbf{\alpha}_v^2}, \\
      \mathbf{v}_2 &= \frac{\mathbf{\alpha}_v}{\alpha_2}.
   \end{align*}

   4. Construct and apply the next orthogonal transformation:
   \begin{align*}
      \rho_1 &= \sqrt{(\bar{\rho_1})^2 + \beta_2^2}, \\
      c_1 &= \frac{\bar{\rho_1}}{\rho_1}, \\
      s_1 &= \frac{\beta_2}{\rho_1}, \\
      \theta_2 &= s_1 \alpha_2, \\
      \bar{\rho_2} &= -c_1 \alpha_2, \\
      \phi_1 &= c_1\bar{\phi_1}, \\
      \bar{\phi_2} &= s_1 \bar{\phi_1}.
   \end{align*}

   5. Update $\mathbf{x}$ and $\mathbf{w}$:
   \begin{align*}
      \mathbf{x}_1 &= \mathbf{x}_0 + \frac{\phi_1}{\rho_1} \mathbf{w}_1, \\
      \mathbf{w}_2 &= \mathbf{v}_2 - \frac{\theta_2}{\rho_1} \mathbf{w}_1.
   \end{align*}

   6. Reset the loop variables:
   \begin{align*}
      \beta_1 &= \beta_2, \\
      \mathbf{u}_1 &= \mathbf{u}_2, \\
      \alpha_1 &= \alpha_2, \\
      \mathbf{v}_1 &= \mathbf{v}_2, \\
      \bar{\rho_1} &= \bar{\rho_2}, \\
      \bar{\phi_1} &= \bar{\phi_2}, \\
      \mathbf{x}_0 &= \mathbf{x}_1, \\
      \mathbf{w}_1 &= \mathbf{w}_2.
   \end{align*}

   7. Check for convergence.

At this point, the authors do not propose a criterion for when we should stop iterating. However, we have chosen to use the criterion that [LSQR in scipy](https://docs.scipy.org/doc/scipy/reference/generated/scipy.sparse.linalg.lsqr.html) [@virtanen2020scipy; @scipy_lsqr] uses. First, compute the residual between the response vector $y$ and the estimate as

$$
r = y-Ax_0
$$

and then we stop iterating when

$$
\|r\| \leq \text{tol}\cdot \|A\| \cdot \|x_0\| + \text{tol} \cdot \|b\|.
$$

`tol` is a parameter for tolerance level of the difference that we accept. In our algorithm, it is by default $1e-6$.

## 1.4. Implementation
Assuming we want to fit regression coefficient vector $\beta$ on the scaled matrix $X$ with response vector $y$. We can compute these estimates using lazy computation. Assuming the original sparse matrix `A`, create an instance of `LazyMatrix` and run the method `lazymatrix::lsqr`.
```{r setup}
library(lazymatrix)
 # 1. Define sparseMatrix
set.seed(123)
n_row <- 50
n_col <- 10
i <- c(
  1:50,
  sample(1:50, 20, replace = TRUE),
  sample(1:50, 15, replace = TRUE)
)
j <- c(
  rep_len(1:10, 50),
  sample(1:10, 20, replace = TRUE),
  sample(1:10, 15, replace = TRUE)
)
pairs <- unique(data.frame(i = i, j = j))
i <- pairs$i
j <- pairs$j
x <- rnorm(length(i))
A <- Matrix::sparseMatrix(i = i, j = j, x = x, dims = c(n_row, n_col))

# 2. Define response vector
y <- rnorm(nrow(A))

# 3. Define LazyMatrix
X <- LazyMatrix(A, "sd", "mean")

# 4. Perform lsqr
lazy_beta <- lsqr(X, y)
```
Now,the non-lazy approach would be to scale `A` using `base::scale`and then use something like `stats::lm.fit` to get the estimates. This would imply creating a copy of `A` with non-zero elements taking up unneccessary memory and being much slower generally.
```{r non lazy approach}
scaled_a <- base::scale(A)
non_lazy_beta <- stats::lm.fit(scaled_a, y)$coefficients
```
The estimates are equivalent as is shown below.
```{r result}
isTRUE(all.equal(as.vector(lazy_beta), as.vector(non_lazy_beta)))
```
# 2. Gradient Descent
## 2.1. Introduction
In modern statistical modeling, the gradient descent algorithm is fundamental for any kind of modern machine learning approach. The approach we propose based on lazy evaluation, is how to perform the simplest form of gradient descent on a scaled matrix. As many of our implementations show, it is sufficient to use a small set of matrix operations in order to perform this algorithm.

## 2.2. Mathematical Background
The algorithm follows the approach as explained by @ng2025cs229. Assume the cost function $J(\theta)$ defined as

$$
J(\theta)=\frac{1}{2}\sum_{i=1}^n(h_\theta(x^{(i)})-y^{(i)})^2
$$

where $h_\theta(x)=\theta^Tx$, $x$ being the elements of the design matrix and $\theta$ the parameters. In gradient descent, we usually talk about two types of parameters, weights and biases. In the notation above, they are both included within the parameter vector $\theta$. This is done by letting $x_0=1$ so that the first term in $\sum_{i=0}^n\theta_ix_i$ includes the additive bias term. $y$ is the response vector, so we recognize $J$ to be the cost function representing the residual sum of squares. Hence, the goal is to choose $\theta$ to minimize $J$. We start with an inistial guess of the parameter and repeatedly update it to make the cost function smaller. This is done with the gradient descent algorithm defined as

$$
\theta_j:=\theta_j - \alpha\frac{\partial}{\partial\theta_j}J(\theta)
$$
where the partial derivative is the derivative of the cost function with regards to the parameter. $\alpha$ is the learning rate that usually takes a small value to control how fast the algorithm converges. Computing this derivative is straight forward and we get the expression

$$
\frac{\partial}{\partial\theta_j}J(\theta)=(h_\theta(x)-y)x_j
$$

providing us with the update rule

$$
\theta_j:=\theta_j+\alpha(y^{(i)}-h_\theta(x^{(i)}))x_j^{(i)}.
$$

## 2.3. Implementation
The following gradient descent algorithm works as helper function for which we will compare the lazy-approach to the non-lazy.
```{r gradient descent algorithm}
gradient_descent_helper <- function(
  x,
  y,
  w_init,
  b_init,
  learning_rate,
  n_epochs
) {
  w <- w_init
  b <- Matrix::Matrix(b_init, nrow = nrow(x))
  n <- nrow(x)
  for (epoch in 1:n_epochs) {
    y_pred <- x %*% w + b
    error <- y_pred - y
    w <- w - (learning_rate * crossprod(x, error)) / n
    b <- b - (learning_rate * sum(error)) / n
  }
  return(list(w = w, b = b))
}
```
Compared to the definition above, in this algorithm we include the bias term as `b`. This results a list of the two parameter vectors, weights and biases. `epochs` is the number of time we choose to iterate. As the result of `LazyMatrix %*% vector` results in a `Matrix`-object, the addition of the bias term can be handled completely within the `Matrix` framework. The key of `lazymatrix` is to handle the heavy matrix operations lazily which is done in this case using `%*%` and `crossprod`. Now, we use the sparse matrix `A` from section #1 and it's `LazyMatrix` `X` and set initial values randomly for the weights and biases and the response vector `y_true`.

```{r initialization}
  set.seed(4567)
  p <- ncol(X)
  w_init <- stats::rnorm(p)
  b_init <- stats::rnorm(1)
  y_true <- stats::rnorm(nrow(A))
  learning_rate <- 0.01
  n_epochs <- 10
```
Note that the function returns the parameters from the iterative algorithm.
```{r}
pars_lazy <- gradient_descent_helper(
   x = X,
   y = y_true,
   w_init = w_init,
   b_init = b_init,
   learning_rate = learning_rate,
   n_epochs = n_epochs
  )
print("First 5 parameter weights and the bias term.")
print(pars_lazy$w[1:5])
print(pars_lazy$b[1])
```
These parameters can now be used for predictions and are comparible with the non-lazy way of using `scale(A)`and then running gradient descent.

## 2.4. Discussion
Nowadays there exists much more complex versions of the gradient descent algorithm, but this example proves that just with a few matrix operations being performed lazily, we can avoid storing heavy copies of large matrices and only perform operations when they are needed. The idea is to implement methods to `lazymatrix` which allows for more complex algorithms such as Stochastic gradient descent.

# 3. Principal Component Analysis
## 3.1. Introduction
A fundamental statistical algorithm for dimension reduction is principal component analysis, or shorter pca. In base R, the function used is `prcomp` which is done by breaking up the matrix `A` into it's singular value decomposition, svd. This implies that we require a lazy implementation of `svd` in order for principal component analysis to work. Moreover, many methods have been proposed to perform SVD on sparse matrices and in `lazymatrix` we have chosen to follow the iterative algorithm proposed by Baglama and Reichel(2005) [@baglama2005augmented] which the authors have also implemented in the R package `irlba`. As this method uses iterative methods, we can easily include `lazymatrix`'s methods for `%*%` and `crossprod` in order to get a working algorithm for svd. However, as the authors themselves say in the paper, the `irlba` function will not compute all singular values and proposes the user to use `base::svd` if one wishes to compute all singular values. This would however involve materializing `A` breaking the lazy computation in the process. Therefore, we work solely with the algorithm proposed by `irlba`.

## 3.2. Mathematical Background
### 3.2.1. Partial Singular Value Decomposition with `irlba`
Assume the matrix $\tilde{A}$ being the scaled version of `A` being $n\times p$. We wish to perform a partial singular value decomposition, where the first step is using a Lanczos process which can be described as

1. Choose a random unit vector $p_1\in R^p$
2. Compute $\tilde{A}p_1=z\in R^n$, transforming $p_1$ from $R^p$ to $R^n$
3. Normalize the vector $q_1=\frac{z}{||z||}$ and let $\alpha_1=||z||$. This builds up the matrix $B$ as it's first diagonal entry.
4. Compute the residual vector by first returning to $R^p$ with $\tilde{A}^Tq_1=z\in R^p$ and let

$$
r=z-\alpha_1p_1 \in R^p.
$$

5. Compute the length of the residual vector $\beta_1=||r||$ and let

$$
p_2=\frac{r}{\beta_1}.
$$

6. Continue to build up $B$ with $\beta_1$ as

$$
B=\begin{pmatrix}
\alpha_1 & \beta_1 &
\end{pmatrix}
$$

At the next iteration, $p_2$ is used as input vector and we start over until we reach convergence after $m$ steps, resulting in the bidiagonal form of $B\in R^{m\times m}$. At every point when matrix multiplication is being computed with $\tilde{A}$ or $\tilde{A}^T$, lazy computation through `lazymatrix` is being implemented.
Once we have performed every iteration in the Lanczos process, we can stack the right basis vectors $p_1, p_2, \ldots, p_m$ into the matrix $P_m\in R^{p\times m}$ and the left basis vectors $q_1, q_2,...,q_m$ into $Q_m\in R^{n\times m}$ resulting in the equations

$$
\begin{align*}
\tilde{A}P_m &= Q_mB_m\\
\tilde{A}^TQ_m &= P_mB_m^T+r_me_m^T
\end{align*}
$$

where $r_m\in R^p$ is the residual vector which is orthogonal to all columns in $P_m$ and $e_m\in R^m$ being the $m$:th standard vector in $R^m$. We start by looking at

$$
\tilde{A}P_m=Q_mB_m
$$

where we know that $B_m$ is bidiagonal and since $m<<\min(n,p)$, this is a smaller matrix and the decomposition can be written as

$$
B_m=U_B\Sigma_B V_B^T
$$
where

$$
\begin{align*}
\Sigma_B&=\text{diag}(\sigma_1^B, \sigma_2^B,\ldots,\sigma_m^B)\newline
U_B&\in R^{m\times m} \\
V_B&\in R^{m\times m}.
\end{align*}
$$

This decompoisition can be used to get approximate solutions to the svd of $\tilde{A}$ with the relationships

$$
\begin{align*}
\tilde{\sigma}_j&=\sigma_j^B\\
\tilde{u}_j &= Q_mu_j^B \in R^n\\
\tilde{v}_j &= P_mv_j^B \in R^p.\\
\end{align*}
$$

This can be validated easily by looking first at the left singular vectors

$$
\tilde{A}\tilde{v}_j=\tilde{A}P_mv_j^B=Q_mB_mv_j^B=\sigma_j^BQ_mu_j^B=\tilde{\sigma}_j\tilde{u}_j
$$

which implies that $(\tilde{\sigma}_j,\tilde{u}_j, \tilde{v}_j)$ satisfies the left singular vector equation. Instead, the right side has a residual error if we look at

$$
\tilde{A}^TQ_m=P_mB_m^T+r_me_m^T
$$

where it can be shown that

$$
\tilde{A}^T\tilde{u_j}=\tilde{\sigma}_j\tilde{v_j}+(e_m^Tu_j^B)r_m.
$$

This error term is proportional to the last computation of $u_j^B$ and residual $r_m$.

Now, we are not neccessarily interested in all $m$ iterations, but it is sufficient to choose $k<m$ as the number of triplets $(\tilde{\sigma_j}, \tilde{u}_j,\tilde{v}_j)$ for which we aim to compute. This is also a parameter in the `svd` call for `lazymatrix`. What we do is after $m$ iterations, we choose the $k$ biggest singular values $\sigma^B_1, \sigma^B_2,\ldots,\sigma^B_k$ and their correspondent left and right vectors. We see whether these have converged based on the criterion

$$
|\beta_m(e_m^Tu_j^B)|<\text{tol}\cdot \tilde{\sigma}_j
$$

and $\beta=||r_m||$ and `tol` being a tolerance parameter. If all $k$ have converged, the algorithm is done. If instead we do not have convergence, we start the Lanczos process again using subspace spanned by the $k$ right singular vectors $\tilde{v}_1, \tilde{v}_2,\ldots, \tilde{v}_k$ and from there we run another $m-k$ runs of the process. This ensures that the previously computed singular values $\tilde{\sigma}_1, \ldots, \tilde{\sigma}_k$ are carried over into the new $B_m$​, rather than restarting from scratch.

### 3.2.2. Principal Component Analysis with Partial SVD
As the regular `stats::prcomp` function uses `base::svd` for dimension reduction, the equivalent method for principal component analysis in `lazymatrix` is identical but with the custom method using partial svd with irlba.

## 3.3. Implementation
As previously, we define a sparse matrix for which we will perform singular value decomposition and thereafter principal component analysis. We use the same sparse matrix from above `A` and use it do define the lazy matrix `X`.

```{r svd, warning=FALSE}
#3. SVD
svd_lazy <- svd(X)

# 4. Print singular values
print("The singular values are: ")
svd_lazy$d
print("The first 6 rows of the left singular vectors are: ")
head(svd_lazy$u)
print("The frist 6 rows of the right singular vectors are: ")
head(svd_lazy$v)
```

This function has been incorporated within the S4 method for the `LazyMatrix` object if we call `prcomp`. Hence, working with the `LazyMatrix` is equivalent to working with a regular matrix object.

```{r prcomp, warning=FALSE}
pca_lazy <- prcomp(X)
print("The first 6 rows of the principal components are: ")
head(pca_lazy$rotation)
```
## 3.4. Conclusion
What we encounter when using iterative methods for performing sparse svd, is that we will not compute all singular values as is the case when calling `base::svd`. One could argue that the method should, if possible, compute all of them using a regular `base::svd` call. However, this breaks sparsity and the lazy structure, materializing the matrix and hence loosing the lazy computation. Since the use case for `lazymatrix` is mainly for working with large sparse matrices, where computation of all singular values may not even be feasible, we recommend not to use this framework for small matrices where sparsity is not an issue and all singular values may be computed using `base::svd`.

# 4. References
